---
title: "sf_package"
author: "Bryan Milstead"
date: "September 18, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(sf)
library(tidyverse)
```

## my notes

* use [as_spatial()] to convert sf object to sp

```{r as_spatial}
library(sf)
nc <- st_read(system.file("shape/nc.shp", package="sf"))

class(nc)

# sf -> sp
nc_sp <- sf:::as_Spatial(nc$geom) # This works

class(nc_sp)
```



## sf (simple features) package

These 3 vignettes are useful
    * [Simple Features for R](https://cran.r-project.org/web/packages/sf/vignettes/sf1.html)
    * [Reading, Writing and Converting Simple Features](https://cran.r-project.org/web/packages/sf/vignettes/sf2.html)
    * [Manipulating Simple Feature Geometries](https://cran.r-project.org/web/packages/sf/vignettes/sf3.html)
    
### Methods for first vignette
    * use [read_sf] to read shapefile and convert to sf object
    * sf object can be used as a data.frame or a geospatial object
    * use [st_geometry] to see the coordinates
    * [plot] shows a facetted map of the attributes
    * use index or field name for a plot of singe attribute
    * find polygons with holes or exterior rings
        * first observation is the exterior ring
        * subsequent observations are holes or additional exterior rings
    * create and display various geometry types
    * use [write_sf] to write the sf object; can specify shapefile or it will guess which drive to use from the name
    * use [st_transform] to change the CRS (can use ESPG code or proj4)

```{r sf1}
nc <- st_read(system.file("shape/nc.shp", package="sf"))
class(nc)
head(nc)
nc_geom<-st_geometry(nc) #same as nc$geom
  nc_geom
  
plot(nc)
plot(nc[1])
plot(nc[1,1], col = 'grey', add = TRUE) #highlight first obs.

#find polygons with holes or exterior rings
(w <- which(sapply(nc_geom, length) > 1))
plot(nc[w,1], col = 2:7)

nc_geom[[4]][[2]][[1]][1:3,]

#geometry types
p <- rbind(c(3.2,4), c(3,4.6), c(3.8,4.4), c(3.5,3.8), c(3.4,3.6), c(3.9,4.5))
(mp <- st_multipoint(p))
## MULTIPOINT (3.2 4, 3 4.6, 3.8 4.4, 3.5 3.8, 3.4 3.6, 3.9 4.5)
s1 <- rbind(c(0,3),c(0,4),c(1,5),c(2,5))
(ls <- st_linestring(s1))
## LINESTRING (0 3, 0 4, 1 5, 2 5)
s2 <- rbind(c(0.2,3), c(0.2,4), c(1,4.8), c(2,4.8))
s3 <- rbind(c(0,4.4), c(0.6,5))
(mls <- st_multilinestring(list(s1,s2,s3)))
## MULTILINESTRING ((0 3, 0 4, 1 5, 2 5), (0.2 3, 0.2 4, 1 4.8, 2 4.8), (0 4.4, 0.6 5))
p1 <- rbind(c(0,0), c(1,0), c(3,2), c(2,4), c(1,4), c(0,0))
p2 <- rbind(c(1,1), c(1,2), c(2,2), c(1,1))
pol <-st_polygon(list(p1,p2))
p3 <- rbind(c(3,0), c(4,0), c(4,1), c(3,1), c(3,0))
p4 <- rbind(c(3.3,0.3), c(3.8,0.3), c(3.8,0.8), c(3.3,0.8), c(3.3,0.3))[5:1,]
p5 <- rbind(c(3,3), c(4,2), c(4,3), c(3,3))
(mpol <- st_multipolygon(list(list(p1,p2), list(p3,p4), list(p5))))
## MULTIPOLYGON (((0 0, 1 0, 3 2, 2 4, 1 4, 0 0), (1 1, 1 2, 2 2, 1 1)), ((3 0, 4 0, 4 1, 3 1, 3 0), (3.3 0.3, 3.3 0.8, 3.8 0.8, 3.8 0.3, 3.3 0.3)), ((3 3, 4 2, 4 3, 3 3)))
(gc <- st_geometrycollection(list(mp, mpol, ls)))
## GEOMETRYCOLLECTION (MULTIPOINT (3.2 4, 3 4.6, 3.8 4.4, 3.5 3.8, 3.4 3.6, 3.9 4.5), MULTIPOLYGON (((0 0, 1 0, 3 2, 2 4, 1 4, 0 0), (1 1, 1 2, 2 2, 1 1)), ((3 0, 4 0, 4 1, 3 1, 3 0), (3.3 0.3, 3.3 0.8, 3.8 0.8, 3.8 0.3, 3.3 0.3)), ((3 3, 4 2, 4 3, 3 3))), LINESTRING (0 3, 0 4, 1 5, 2 5))

dpar<-par()
par(mfrow=c(2,3))
plot(mp, main = "multiploint")
  box()
plot(ls, main = "linestring")
  box()
plot(mls, main = "multilinestring")
  box()
plot(pol, main = "polygon")
  box()
plot(mpol, main = "multipolygon")
  box()
plot(gc, main = "geometrycollection")
  box()
par<-dpar

#write shapefile
write_sf(nc, "c:/bryan/bryan_temp/nc.shp") # silently overwrites

```



```{r sf2}
b0 = st_polygon(list(rbind(c(-1,-1), c(1,-1), c(1,1), c(-1,1), c(-1,-1))))
b1 = b0 + 2
b2 = b0 + c(-0.2, 2)
x = st_sfc(b0, b1, b2)
a0 = b0 * 0.8
a1 = a0 * 0.5 + c(2, 0.7)
a2 = a0 + 1
a3 = b0 * 0.5 + c(2, -0.5)
y = st_sfc(a0,a1,a2,a3)
plot(x, border = 'red')
plot(y, border = 'green', add = TRUE)

u<-st_union(x)
v<-st_union(y)
plot(u,add=TRUE,border='blue')
plot(v,add=TRUE,border='grey')


plot(st_intersection(st_union(x),st_union(y)), add = TRUE, col = 'red')

par(mfrow=c(2,2), mar = c(0,0,1,0))
plot(x, col = '#ff333388'); 
plot(y, add=TRUE, col='#33ff3388')
title("x: red, y: green")
plot(x, border = 'grey')
plot(st_difference(st_union(x),st_union(y)), col = 'lightblue', add = TRUE)
title("difference(x,y)")
plot(x, border = 'grey')
plot(st_difference(st_union(y),st_union(x)), col = 'lightblue', add = TRUE)
title("difference(y,x)")
plot(x, border = 'grey')
plot(st_sym_difference(st_union(y),st_union(x)), col = 'lightblue', add = TRUE)
title("sym_difference(x,y)")


```

```{r lakes}
#get lakes database
ma<-st_read('Y:/data/lakeMorphometry/LakeMorphGdb.gdb',layer='MidAtlantic02')
ma_g<-st_geometry(ma) #same as nc$geom
w <- which(sapply(ma_g, length) > 1)
w  #no multipart polygons in the database  

ma_g[[1794]] #lake placcid has some large holes (islands)

length(ma_g[[1794]]) # ==1 how are the holes coded?

plot(ma[1794,1])


#get the buffer
placcid<-ma[1794,]
placcid_buf<-st_buffer(placcid,2000)
placcid_dif<-st_sym_difference(st_union(placcid), st_union(placcid_buf))
  
  
  
plot(placcid_buf[1], col = 'red')
plot(placcid[1],add=TRUE, col = 'blue')
plot(placcid_dif[1],add=TRUE, col = 'green')

plot(placcid_dif[1], col = 'green')








# simple lake
lake<-st_polygon(list(rbind(c(0,0), c(1,0), c(3,2), c(2,4), c(1,4), c(0,0))))


# lake with Island
p1 <- rbind(c(0,0), c(1,0), c(3,2), c(2,4), c(1,4), c(0,0))
p2 <- rbind(c(1,1), c(1,2), c(2,2), c(1,1))
island <-st_polygon(list(p1,p2))

#connected lakes
p1 <- rbind(c(0,0), c(1,0), c(3,2), c(2,4), c(1,4), c(0,0))
p2 <- rbind(c(3,2), c(4,3), c(5,3), c(6,2), c(4.5,1), c(3,2))
clakes <-st_polygon(list(p1,p2))

# two lakes
p1 <- rbind(c(0,0), c(1,0), c(3,2), c(2,4), c(1,4), c(0,0))
p2 <- rbind(c(3.5,2), c(4,3), c(5,3), c(6,2), c(4.5,1), c(3.5,2))
tlakes <-st_polygon(list(p1,p2))

par(mfrow=c(2,2))
plot(lake)
plot(island)
plot(clakes)
plot(tlakes)
par(mfrow=c(1,1))


par(mfrow=c(2,2))
plot(st_buffer(lake,2), col = 'red')
plot(lake,add=TRUE, col = 'blue')

plot(st_buffer(island,2))
plot(island,add=TRUE)

plot(st_buffer(clakes,2))
plot(clakes,add=TRUE)

plot(st_buffer(tlakes,2))
plot(tlakes,add=TRUE)

par<-dpar


plot(st_buffer(st_combine(tlakes),2), col = 'red')
plot(tlakes,add=TRUE, col = 'blue')


head(island)

nc <- st_read(system.file("shape/nc.shp", package="sf"))
plot(nc[1])

print(nc[8:15], n = 3)

nc[1:3,9:15]

plot(nc[4,1])
st_geometry(island)

nc_geom <- st_geometry(nc)
nc_geom[[4]][[2]][[1]]
```

