---
title: "tidycensus_vignettes"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.width = 8)
```

### Basic usage of tidycensus Vignette. source: https://github.com/walkerke/tidycensus/blob/master/vignettes/basic-usage.Rmd

To get started working with __tidycensus__, users should load the package along with the __tidyverse__ package, and set their Census API key.  A key can be obtained from <http://api.census.gov/data/key_signup.html>.  

```{r, eval = TRUE}
library(tidycensus)
library(tidyverse)

census_api_key("eb5e651aa616016f0ab03c0a8154eeec47e8c96d")  

#OR

##https://walkerke.github.io/tidycensus/articles/basic-usage.html
##save the census api key:
#cat("CENSUS_API_KEY=eb5e651aa616016f0ab03c0a8154eeec47e8c96d\n",file=file.path(normalizePath("~/"), ".Renviron"),append=TRUE)
##restart R
```

There are two major functions implemented in __tidycensus__: `get_decennial`, which grants access to the 1990, 2000, and 2010 decennial US Census APIs, and `get_acs`, which grants access to the 5-year American Community Survey APIs.  In this basic example, let's look at median gross rent by state in 1990: 

```{r}
m90 <- get_decennial(geography = "state", variables = "H043A001", year = 1990)

head(m90)
```

The function returns a tibble with four columns by default: `GEOID`, which is an identifier for the geographical unit associated with the row; `NAME`, which is a descriptive name of the geographical unit; `variable`, which is the Census variable represented in the row; and `value`, which is the value of the variable for that unit.  By default, __tidycensus__ functions return tidy data frames in which rows represent unit-variable combinations; for a wide data frame with Census variable names in the columns, set `output = "wide"` in the function call.  

As the function has returned a tidy object, we can visualize it quickly with __ggplot2__: 

```{r, fig.height = 8}
m90 %>%
  ggplot(aes(x = value, y = reorder(NAME, value))) + 
  geom_point()
```

## Searching for variables

Getting variables from the Census or ACS requires knowing the variable ID - and there are thousands of these IDs across the different Census files.  To rapidly search for variables, use the `load_variables` function.  The function takes two required arguments: the year of the Census or endyear of the ACS sample, and the dataset - one of `"sf1"`, `"sf3"`, or `"acs5"`. For ideal functionality, I recommend assigning the result of this function to a variable, setting `cache = TRUE` to store the result on your computer for future access, and using the `View` function in RStudio to interactively browse for variables.  

```{r, eval = FALSE}
sf1_2010 <- load_variables(2010, "sf1", cache = TRUE)
sf3_2000 <- load_variables(2000, "sf3", cache = TRUE)
sf3_1990 <- load_variables(1990, "sf3", cache = TRUE)

View(sf1_2010)
View(sf3_2000)
View(sf3_1990)
```

<img src=img/view.png style="width: 100%">

By filtering for "median age" I can quickly view the variable IDs that correspond to my query.  

## Working with ACS data

American Community Survey data differ from decennial Census data in that ACS data are based on an annual sample of approximately 3 million households, rather than a more complete enumeration of the US population.  In turn, ACS data points are __estimates__ characterized by a __margin of error__.  __tidycensus__ will always return the estimate and margin of error together for any requested variables.  In turn, when requesting ACS data with __tidycensus__, it is not necessary to specify the `"E"` or `"M"` suffix for a variable name.  Let's fetch median household income data from the 2011-2015 ACS for counties in Vermont; the endyear is not necessary here as the function defaults to 2015. 

```{r}
vt <- get_acs(geography = "county", variables = "B19013_001", state = "VT")

head(vt)
```

The output is similar to a call to `get_decennial`, but instead of a `value` column, `get_acs` returns `estimate` and `moe` columns for the ACS estimate and margin of error, respectively.  `moe` represents the default 90 percent confidence level around the estimate; this can be changed to 95 or 99 percent with the `moe_level` parameter in `get_acs` if desired. 

As we have the margin of error, we can visualize the uncertainty around the estimate: 

```{r}
vt %>%
  mutate(NAME = gsub(" County, Vermont", "", NAME)) %>%
  ggplot(aes(x = estimate, y = reorder(NAME, estimate))) +
  geom_errorbarh(aes(xmin = estimate - moe, xmax = estimate + moe)) +
  geom_point(color = "red", size = 3) +
  labs(title = "Household income by county in Vermont",
       subtitle = "2011-2015 American Community Survey",
       y = "",
       x = "ACS estimate (bars represent margin of error)")
```

### Spatial data in tidycensus Vignette. Source: https://github.com/walkerke/tidycensus/blob/master/vignettes/spatial-data.Rmd

If requested, __tidycensus__ can return simple feature geometry for geographic units along with variables from the decennial US Census or American Community survey.  By setting `geometry = TRUE` in a __tidycensus__ function call, __tidycensus__ will use the __tigris__ package to retrieve the corresponding geographic dataset from the US Census Bureau and pre-merge it with the tabular data obtained from the Census API.  The following example shows median household income from the 2011-2015 ACS for Census tracts in Orange County, California: 

```{r}
library(tidycensus)
library(tidyverse)
options(tigris_use_cache = TRUE)

orange <- get_acs(state = "CA", county = "Orange", geography = "tract", 
                  variables = "B19013_001", geometry = TRUE)

head(orange)
```

Our object `orange` looks much like the basic __tidycensus__ output, but with a `geometry` list-column describing the geometry of each feature, using the geographic coordinate system NAD 1983 (EPSG: 4269) which is the default for Census shapefiles.  __tidycensus__ uses the Census [cartographic boundary shapefiles](https://www.census.gov/geo/maps-data/data/tiger-cart-boundary.html) for faster processing; if you prefer the TIGER/Line shapefiles, set `cb = FALSE` in the function call. 

As the dataset is in a tidy format, it can be quickly visualized with the `geom_sf` functionality currently in the development version of __ggplot2__: 

```{r}
library(viridis)

orange %>%
  ggplot(aes(fill = estimate, color = estimate)) + 
  geom_sf() + 
  coord_sf(crs = 26911) + 
  scale_fill_viridis(option = "magma") + 
  scale_color_viridis(option = "magma")
```

## Faceted mapping

One of the most powerful features of __ggplot2__ is its support for small multiples, which works very well with the tidy data format returned by __tidycensus__.  Many Census and ACS variables return _counts_, however, which are generally inappropriate for choropleth mapping.  In turn, `get_decennial` and `get_acs` have an optional argument, `summary_var`, that can work as a multi-group denominator when appropriate.  Let's use the following example of the racial geography of Harris County, Texas.  First, we'll request data for non-Hispanic whites, non-Hispanic blacks, non-Hispanic Asians, and Hispanics by Census tract for the 2010 Census, and specify total population as the summary variable.  `year` is not necessary here as the default is 2010.   

```{r}
racevars <- c("P0050003", "P0050004", "P0050006", "P0040003")

harris <- get_decennial(geography = "tract", variables = racevars, 
                  state = "TX", county = "Harris County", geometry = TRUE,
                  summary_var = "P0010001") 

head(harris)
```

We notice that there are four entries for each Census tract, with each entry representing one of our requested variables. The `summary_value` column represents the value of the summary variable, which is total population in this instance.  When a summary variable is specified in `get_acs`, both `summary_est` and `summary_moe` columns will be returned.  

With this information, we can set up an analysis pipeline in which we calculate a new percent-of-total column; recode the Census variable names into more intuitive labels; and visualize the result for each group in a faceted plot.  

```{r}
library(forcats)

harris %>%
  mutate(pct = 100 * (value / summary_value),
         variable = fct_recode(variable,
                               White = "P0050003",
                               Black = "P0050004",
                               Asian = "P0050006",
                               Hispanic = "P0040003")) %>%
  ggplot(aes(fill = pct, color = pct)) +
  facet_wrap(~variable) +
  geom_sf() +
  coord_sf(crs = 26915) + 
  scale_fill_viridis() +
  scale_color_viridis()
```

Beyond this, you might be interested in writing your dataset to a shapefile or GeoJSON for use in external GIS or visualization applications.  You can accomplish this with the `st_write` function in the __sf__ package: 

```{r, eval = FALSE}
library(sf)
st_write(orange, "orange.shp")
```

Your __tidycensus__-obtained dataset can now be used in ArcGIS, QGIS, Tableau, or any other application that reads shapefiles.  

There is a lot more you can do with the spatial functionality in __tidycensus__, including more sophisticated visualization and spatial analysis; look for updates [on my blog](https://walkerke.github.io) and in this space.             
  
  


